---
title: "Phase 1: Comparative Genomics"
author: "Geoffrey Ochieng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(ggplot2)
```

# Phase 1: Comparative Genomics

## Biological Question

Why does *Trypanosoma equiperdum* OVI retain kinetoplast DNA despite belonging to a lineage in which kDNA loss has evolved repeatedly?

Phase 1 investigates whether ATPase γ and other mitochondrial proteins show signatures of functional constraint or compensatory evolution across dyskinetoplastic and akinetoplastic *Trypanozoon* lineages.



## Objectives

- [x] Retrieve sequences for ATPase γ and related genes
- [x] Reconstruct ATPase γ sequences from short-reads datasets
- [x] Generate protein alignments
- [x] Generate multiple sequence alignments
- [ ] Calculate dN/dS (omega) values
- [ ] Compare selection patterns among lineages
- [ ] Test for purifying selection

# Data

*To be added during Phase 1*

## Study Isolates

| Strain | Species | Phenotype |
|----------|----------|----------|
| TREU927 | *T. brucei* | Reference |
| OVI | *T. equiperdum* | Dyskinetoplastic |
| BoTat | *T. equiperdum* | Akinetoplastic |
| IVM-t1 | *T. equiperdum* | Dyskinetoplastic |
| STIB805 | *T. evansi* | Akinetoplastic |
| MU09 | *T. evansi* Type A | Akinetoplastic |
| MU10 | *T. evansi* Type B | Akinetoplastic |
| RoTat 1.2 | *T. evansi* Type A | Dyskinetoplastic |

# Methods

*To be added during Phase 1*

## ATPase γ Extraction

ATPase γ coding sequences were extracted from assembled genomes using BLAST searches against the *T. brucei* ATPase γ reference sequence.

For isolates lacking genome assemblies, ATPase γ sequences were reconstructed by mapping sequencing reads against the OVI ATPase γ coding sequence.

### Read Mapping Pipeline

```bash
minimap2
samtools sort
samtools index
bcftools mpileup
bcftools call
bcftools consensus
transeq
```

## Multiple Sequence Alignment

Protein sequences were translated using EMBOSS `transeq` and aligned using MAFFT v7.

```bash
mafft --auto ATPase_gamma_all.faa > ATPase_gamma_all_aligned.faa
```

# Results

## ATPase γ Mutation Validation

ATPase γ sequences were recovered for all study isolates.

### Mutation Summary

| Strain | ATPase γ status |
|----------|----------|
| TREU927 | Wild type |
| OVI | Wild type |
| IVM-t1 | Wild type |
| STIB805 | A281del |
| MU09 | A281del |
| RoTat | A281del (heterozygous) |
| MU10 | M282L (heterozygous) |
| BoTat | A273P |

## Validation of RoTat ATPase γ

Initial haploid consensus calling suggested a wild-type ATPase γ sequence.

Re-analysis using diploid variant calling recovered a heterozygous deletion:

| Position | REF | ALT | Genotype |
|----------|----------|----------|----------|
| 838 | TCTGCT | TCT | 0/1 |

This deletion removes nucleotides 841–843 (GCT), corresponding to the canonical A281del mutation previously reported in Type A *T. evansi* isolates.

## Validation of MU10 ATPase γ

Initial haploid consensus reconstruction yielded a wild-type methionine residue at codon 282.

Inspection of the diploid VCF revealed:

| Position | REF | ALT | Genotype |
|----------|----------|----------|----------|
| 844 | A | T | 0/1 |

The A844T polymorphism corresponds to the M282L ATPase γ mutation reported in Type B *T. evansi* isolates.

## Methodological Observation

Haploid consensus generation masked biologically relevant heterozygous ATPase γ variants in both MU10 and RoTat.

Direct inspection of BAM files, pileup data, and diploid VCFs was necessary to recover the expected compensatory mutations.

# Preliminary Interpretation

The ATPase γ dataset recapitulates the major compensatory mutation classes reported in dyskinetoplastic and akinetoplastic *Trypanozoon* lineages:

- Wild type (TREU927, OVI, IVM-t1)
- A281del (STIB805, MU09, RoTat)
- M282L (MU10)
- A273P (BoTat)

These results validate the sequence reconstruction workflow and provide the foundation for downstream codon-based selection analyses.

### Pipeline Validation

This analysis demonstrated that haploid consensus generation can obscure biologically relevant heterozygous mutations. Validation through direct inspection of BAM, pileup, and diploid VCF data was essential for accurate mutation characterization.


## IQ-TREE statistics
ATPase γ was highly conserved across all examined strains. Of 915 aligned nucleotide positions, 900 (98.4%) were invariant and only six sites were parsimony informative. This extensive conservation is consistent with strong evolutionary constraint on ATPase γ despite differences in kinetoplast DNA status.

## ATPase γ Dyskinetoplasty-Associated Residue Analysis

Four ATPase γ residues previously associated with dyskinetoplasty
(L262, A273, A281, and M282) were examined across the eight taxa.
Residues were extracted from the final ATPase γ protein alignment
using *T. brucei* TREU927 as the reference sequence.

```{r mutation-matrix-import, message=FALSE}

library(tidyverse)
library(gt)

mutation_matrix <- read_csv(
  "01_comparative_genomics/results/alignment/ATPase_gamma_mutation_matrix.csv"
)

mutation_matrix

```r
```{r mutation-matrix-table}

mutation_table <- mutation_matrix %>%
  gt() %>%
  tab_header(
    title = "ATPase γ Mutation Matrix",
    subtitle = paste(
      "Dyskinetoplasty-associated residues L262, A273, A281 and M282"
    )
  ) %>%
  cols_label(
    Strain = "Strain",
    `262` = "L262",
    `273` = "A273",
    `281` = "A281",
    `282` = "M282"
  ) %>%
  tab_source_note(
    source_note =
      "* T. evansi MU10 carries a heterozygous M282L variant "
      "detected by VCF analysis (GT=0/1; AD=45,32; DP=81). "
      "The consensus sequence used for phylogenetic reconstruction "
      "retained methionine at this position."
  )

mutation_table

Immediately below the table:

```
markdown
### Interpretation

The ATPase γ mutation analysis revealed that *T. equiperdum* OVI retained
the canonical amino acid state at all four dyskinetoplasty-associated
positions. OVI was identical to the kDNA-dependent reference strain
*T. brucei* TREU927.

In contrast, dyskinetoplastic lineages showed lineage-specific variation:
*T. equiperdum* BoTat carried A273P, *T. evansi* MU09 and STIB805 carried
A281 deletions, and *T. evansi* MU10 contained a heterozygous M282L variant.

These observations indicate that OVI lacks known ATPase γ adaptations
associated with dyskinetoplasty and retains a sequence consistent with
continued mitochondrial constraint.

## Supplementary Metadata

The ATPase γ comparative dataset consisted of eight Trypanozoon taxa.
Strain information, kDNA phenotype classification, sequence source,
and accession/read identifiers are provided in Supplementary Table S1.

```{r supplementary-metadata}

library(readr)
library(gt)

metadata <- read_csv(
  "01_comparative_genomics/data/metadata/ATPase_gamma_metadata.csv"
)

metadata %>%
  gt() %>%
  tab_header(
    title = "Supplementary Table S1",
    subtitle = "Metadata for ATPase γ comparative genomics dataset"
  )

```

## Suggested Supplementary Table caption

**Supplementary Table S1. Metadata for ATPase γ comparative genomics dataset.**  
Information includes strain designation, species identity, kinetoplast DNA phenotype,
sequence source, accession or sequencing read identifier, and reconstruction method. 
Genome-derived sequences were extracted from available assemblies, whereas strains without 
suitable assemblies were recovered using reference-guided consensus calling from raw sequencing 
reads.

---

One additional benefit: when we move to the next phase (POLIB, POLIC, POLID, TAC genes), 
this same metadata table can become the **master sample sheet** for the entire project. 
We can simply add columns for each gene analysis rather than creating new metadata files 
repeatedly. This will make the final manuscript workflow much cleaner.

# Next Steps

- Generate codon alignment
- Construct ATPase γ phylogeny
- Estimate pairwise dN/dS
- Run codeml branch analyses
- Compare ATPase γ evolutionary constraint among lineages

